---
title: "Activity Analysis - Reproducible Research Assignment 1"
author: "Sonia Tripolitano"
date: "26 October 2017"
output: html_document
---

## Read in data, examine and process

```{r setwd, echo=FALSE}

setwd("C:/Users/Stephane/Documents/Sonia/Learning/coursera/4_Repro_Research/Data")
# If this is in a seperate code chunk to the read.csv the .Rmd document
# must be saved in the same place as the data or it didn't work
```

```{r readdat, echo=TRUE}

active<-read.csv("activity.csv")
active$date<-as.Date(active$date)
str(active)
summary(active)
head(active)
```

Create a summary dataset which sums up the steps by day

```{r summarise}
summ<-data.frame(tapply(active$steps,active$date,sum))
daily<-data.frame(date=as.Date(names(summ[,1])),steps=as.numeric(summ[1]$tapply.active.steps..active.date..sum.))
str(daily)
summary(daily)
```

## Steps taken per day and report mean and median
```{r hist, echo=FALSE}

hist(daily$steps,breaks=20,col="blue",main="Histogram of Daily Steps",
     xlab="Total Steps",ylab="Number of Days")

```

```{r meanmed, echo=FALSE}
mean1<-as.integer(mean(na.exclude(daily$steps)))

median1<-as.integer(median(na.exclude(daily$steps)))
```
*The mean number of steps per day is `r mean1` and the median is `r median1`*

## Daily Activity Pattern
### Summarise data to find average steps per interval
````{r summariseb}
active2<-active[complete.cases(active),] # Remove NAs
summ2<-data.frame(tapply(active2$steps,active2$interval,mean)) 
intervals<-data.frame(intervals=as.integer(names(summ2[,1])),steps=as.numeric
                      (summ2[1]$tapply.active2.steps..active2.interval..mean.))
summary(intervals)
```

```{r plot2, echo=FALSE}
plot(x=intervals$intervals,y=intervals$steps,type="l",
     ylab="Average Steps Taken",xlab="Interval",main="Time Series of Average Steps",col="green")

int1<-as.integer(intervals[intervals$steps==max(intervals$steps),1])
```

*The 5-minute interval which contains the maximum number of steps is `r int1`*

## Imputing missing values

I have established previously that only the "steps" column contains NAs.
```{r impute1, echo=FALSE}
x<-as.integer(nrow(active[complete.cases(active)==F,]))
```
*There are `r x` number of rows with missing values*

### Impute missing values using average steps per interval 
```{r impute2}
names(intervals)<-c("interval","avgsteps") # Change names suitable for merging datasets
active3<-merge(active,intervals,by="interval") 
active3$stepsfilled<-active3$steps # Use average steps per interval for missing values
active3[is.na(active3$steps),]$stepsfilled<-active3[is.na(active3$steps),]$avgsteps
head(active3,20) # Check that imputation is correct

active4<-active3[,c(1,3,5)]
names(active4)[3]<-c("steps")

# Summarise to get the steps per day
summ3<-data.frame(tapply(active4$steps,active4$date,sum))
dailyb<-data.frame(date=as.Date(names(summ3[,1])),steps=as.numeric(summ3[1]$tapply.active4.steps..active4.date..sum.))
```

```{r plot3, echo=FALSE}
hist(dailyb$steps,breaks=20,col="yellow",main="Histogram of Daily Steps with Imputation", xlab="Total Steps",ylab="Number of Days")
mean2<-as.integer(mean(dailyb$steps))
med2<-as.integer(median(dailyb$steps))
```

*The mean and median total number of steps taken per day is `r mean2` and `r med2`*

The mean does not differ from the original since the imputation was done using mean.
The median changes slightly.

## Activity pattern weekdays vs weekends
Create a new dataset with a weekend/ weekday flag
```{r}
active4$weekday<-weekdays(active4$date)
active4$weekdayflag<-"Weekday"
active4[active4$weekday==c("Sunday","Saturday"),]$weekdayflag<-"Weekend"
head(active4) # Check that flag is correct
```

After summarising data to get the steps per interval per weekend/ weekday flag,
create panel plot of Time Series of Avg Daily Steps taken
```{r echo=FALSE}
summ4<-data.frame(tapply(active4[active4$weekdayflag=="Weekday",]$steps,active4[active4$weekdayflag=="Weekday",]$interval,mean))
names(summ4)<-"average"
summ4$flag<-"Weekday"

summ4b<-data.frame(tapply(active4[active4$weekdayflag=="Weekend",]$steps,active4[active4$weekdayflag=="Weekend",]$interval,mean))
names(summ4b)<-"average"
summ4b$flag<-"Weekend"

intervals2<-data.frame(intervals=as.integer(names(summ4[,1])),steps=as.numeric
                       (summ4[1]$average),flag=summ4[2])
intervals2b<-data.frame(intervals=as.integer(names(summ4b[,1])),steps=as.numeric
                       (summ4b[1]$average),flag=summ4b[2])

intervalsall<-rbind(intervals2,intervals2b)

library(lattice)
xyplot(intervalsall$steps~intervalsall$intervals|intervalsall$flag, type="l",
            main="Avg Steps Taken by Weekday/ Weekend",
            xlab="Interval", ylab="Avg Steps Taken", cex=1,
            layout=c(1,2))
```
